Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife

### 2023-04-25 analysis/paper/paper.Rmd
Add descrip of dif btw algoritms and comp and mem costs

<!-- TO DO: number analysis scripts to clarify workflow -->

@@

The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems. For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)). The simple underlying grid is less computationally costly than methods designed to support curvature or switchbacks required in operational road design [@stuckelberger_improved_2007]. The algorithm starts with the landing nearest to an existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing. The method was developed by Anderson and Nelson [-@anderson_projecting_2004] who noted that the projected network is sensitive to the ordering of targets and recommended a heuristic for connecting the targets and finding optimal branch locations to deal with this sensitivity.

@@

TO DO: talk about key differences from Hardy algorithm.
TO DO: talk about implications of algorithms for speed/computational cost.

>>>>

@@

The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems (STAPs). For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)). The simple underlying grid is less computationally costly than methods designed to support curvature or switchbacks required in operational road design [@stuckelberger_improved_2007]. The algorithm starts with the landing nearest to an existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing. The method was developed by Anderson and Nelson [-@anderson_projecting_2004] who noted that the projected network is sensitive to the ordering of targets and recommended a heuristic for connecting the targets and finding optimal branch locations to deal with this sensitivity.

@@

Hardy's method [-@hardy_forest_2019;-@hardy_landis-ii_2023] is similar to our DLCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the distance from the existing road network. One difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network. On the other hand, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two. However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network. It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network. Hardy's method also has a different approach for building roads inside cutblocks, it iterates over each cell in the harvest area and builds a road to it if it is not within the skidding distance of an existing road. In our algorithm we create landing points in the harvested area to serve as the starting points for the roads to be built based on an expected landing density.

<!--
TO DO: should this paragraph about speed and memory go in the results or discussion rather than methods?
-->
The MST method is expected to be closer to an optimal solution than either of the LCP methods described above but at the cost of being more computationally intensive and taking longer to solve. The Hardy QGIS plugin was much faster to run than our R package and more robust to analyzing large landscapes. The `igraph` package that we used to implement the Dijkstra algorithm requires the values of the cost raster to be read into memory in R which can lead to high memory usage on larger landscapes. See Appendix X for a more detailed exploration of the speed and memory requirements of the different methods.
